### 2019-09-16 ic-journal/ccs.tex
Removed all reviewed highlights in the first part of the paper; re-enable highlights

\hl{Here we recall the core theory of CCS.}
We assume a possibly infinite set of names $\mathscr{L} = \{a, b,
\ldots\}$ \hl{yielding} input and $\overline{\mbox{output}}$ labels (actions), plus a special invisible
action $\tau \notin \mathscr{L}$, and \hl{another possibly infinite
  set of agent} variables $A, B,
\ldots \in \mathscr{X}$.
The class of CCS processes is then inductively
defined from $\nil$ \hl{(the terminated process)} and agent variables by the operators
of \emph{prefixing} (.), \emph{parallel composition} ($|$),
\emph{summation}
(or \emph{binary choice}, $+$),
\emph{restriction} ($\nu$), \emph{relabeling} ($[\cdot]$) and
\emph{recursion} ($\texttt{rec}$):
\begin{equation*}
\begin{array}{cccl}
\mu  & := & \tau\!\!\!\! & \midd \; a  \; \midd \;  \outC a  \\
P  & := & \nil\!\!\!\! & \midd \;  \mu . P \; \midd \;  P_1 |  P_2 \; \midd  \;
P_1 + P_2 \; \midd 
(\res L)\, P
\; \midd \; P\; [\mathit{rf}]  % relabelling
\;  \midd \;  A \; \midd \; \recu A  P
\end{array}
\end{equation*}
\hl{In our presentation of CCS, 
the restriction operator takes a set of  labels $L
    \subseteq \mathscr{L}$, rather than a single one.  
The relabeling operator takes a relabeling function
    $\mathit{rf} \colon \mathscr{L} \cup \overline{\mathscr{L}} \cup
    \{\tau\} \rightarrow \mathscr{L} \cup \overline{\mathscr{L}} \cup
    \{\tau\}$, that can  handle multiple actions including $\tau$. A valid
    relabeling function $\mathit{rf}$ must however 
satisfy $\mathit{rf}(\tau) =
    \tau$ and $\forall l\in\mathscr{L} \cup \overline{\mathscr{L}}.\, \mathit{rf}(\overline{l}) =
    \overline{\mathit{rf}(l)}$ (with $\bar{\bar l} = l$ for all $l \in
    \mathscr{L}$).}
% \hl{Notice the (minor) differences with other CCS literature
% regarding the above definition:} 
% \begin{enumerate}
%   \item Some literature adopts possibly infinite sums over an index
%     set ($\sum_{i\in I} P_i$) while we only consider binary (thus
%     finite) sums, which is less expressive.
%   \item Our restriction operator takes a set of (input) labels $L
%     \subseteq \mathscr{L}$, while some CCS literature only restricts
%     one label (for each restriction operator);
%   \item Our relabeling operator takes a relabeling function
%     $\mathit{rf} \colon \mathscr{L} \cup \overline{\mathscr{L}} \cup
%     \{\tau\} \rightarrow \mathscr{L} \cup \overline{\mathscr{L}} \cup
%     \{\tau\}$, which
%     is capable to handle multiple actions including $\tau$. A valid
%     relabeling function $\mathit{rf}$ must satisfy $\mathit{rf}(\tau) =
%     \tau$ and $\forall l\in\mathscr{L} \cup
%     \overline{\mathscr{L}}.\, \mathit{rf}(\overline{l}) = 
%     \overline{\mathit{rf}(l)}$ (with $\bar{\bar l} = l$ for all $l \in
%     \mathscr{L}$); Same expressiveness
%     with single-label relabeling.
%   \item The recursion operator $\texttt{rec}$ has the same
%   expressiveness with  
%     \emph{process constants} adopted in some literature. For instance, a
%     CCS process $a.C$ with the constant $C \DSdefi b.C$ can be
%     expressed as $a.(\texttt{rec}\, C. (b.C))$.
% \end{enumerate}
We sometimes omit \hl{a} trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil$.
A CCS process $P$ may \hl{evolve} to another one, say $P'$, under
  an action $\mu$, written by $P \arr{\mu} P'$.
The \hl{transition semantics} of CCS processes is given by means of
a Labeled Transition System (LTS) \hl{expressed in Structural Operational
Semantics (SOS) rules} shown in Fig.~\ref{f:LTSCCS}.
(The symmetric versions of the \hl{SOS} rules for
parallel composition and binary choice are omitted.)
A CCS processes uses only \emph{guarded sums} if all occurrences of
sums are in the form $a.P + b.Q$. 
% \hl{(Otherwise we may call the sum
% a \emph{direct} one, e.g. $P + Q$ if either $P$ or $Q$ are not in the
% form of prefixes.)}
 The \emph{immediate derivatives} of a
process $P$ are elements of the set $\{P' \st P \arr\mu P' \mbox{
  for some $\mu$}\}$.
We use $\ell$ to range over
visible actions (i.e.~inputs or outputs, excluding  $\tau$)
and $\alpha, \mu$  to range over all actions.
\begin{figure*}[t]
\begin{center}
\vskip .1cm
 $\displaystyle{  \over  \mu.  P \arr\mu
P } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P + Q   \arr\mu
P'  } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
P' | Q } $  $ \hb$   
\hskip .3cm
  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
\arr{\outC a }Q'  \over  P|  Q \arr{ \tau} P'
|  Q'  }$ 
\\
\vspace{.2cm}
$\displaystyle{ P \arr{\mu}P' \over
 (\res L\!)\, P \arr{\mu} (\res L\!)\, P'} $ $ \mu, \outC\mu \not\in L$
$ \hb$
%
$\displaystyle{ P \sub {\recu A P} A \arr{\mu}P' \over
 \recu A P   \arr{ \mu} P'  } $
\hskip .5cm  
$\displaystyle{ P \arr{\mu} P' \over
 P \;[r\!f] \arr{r\!f(\mu)} P' \;[r\!f]} $
$ \hb$ %  &
\end{center}
\caption{Structural Operational Semantics of CCS}
\label{f:LTSCCS}
\end{figure*}

>>>>

Here we recall the core theory of CCS.
We assume a possibly infinite set of names $\mathscr{L} = \{a, b,
\ldots\}$ yielding input and $\overline{\mbox{output}}$ labels (actions), plus a special invisible
action $\tau \notin \mathscr{L}$, and another possibly infinite
set of agent variables $A, B, \ldots \in \mathscr{X}$.
The class of \hl{all} CCS \hl{terms (or expressions)} is then inductively
defined from $\nil$ (the terminated process) and agent variables by the operators
of \emph{prefixing} (.), \emph{parallel composition} ($|$),
\emph{summation}
(or \emph{binary choice}, $+$),
\emph{restriction} ($\nu$), \emph{relabeling} ($[\cdot]$) and
\emph{recursion} ($\texttt{rec}$):
\begin{equation*}
\begin{array}{cccl}
\mu  & := & \tau\!\!\!\! & \midd \; a  \; \midd \;  \outC a  \\
P  & := & \nil\!\!\!\! & \midd \;  \mu . P \; \midd \;  P_1 |  P_2 \; \midd  \;
P_1 + P_2 \; \midd 
(\res L)\, P
\; \midd \; P\; [\mathit{rf}]  % relabelling
\;  \midd \;  A \; \midd \; \recu A  P
\end{array}
\end{equation*}
In our presentation of CCS, the restriction operator takes a set of labels $L
  \subseteq \mathscr{L}$, rather than a single one.  
The relabeling operator takes a relabeling function
  $\mathit{rf} \colon \mathscr{L} \cup \overline{\mathscr{L}} \cup
    \{\tau\} \rightarrow \mathscr{L} \cup \overline{\mathscr{L}} \cup
    \{\tau\}$, that can  handle multiple actions including $\tau$. A valid
    relabeling function $\mathit{rf}$ must however 
satisfy $\mathit{rf}(\tau) =
    \tau$ and $\forall l\in\mathscr{L} \cup \overline{\mathscr{L}}.\, \mathit{rf}(\overline{l}) =
    \overline{\mathit{rf}(l)}$ (with $\bar{\bar l} = l$ for all $l \in
    \mathscr{L}$).
We sometimes omit a trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil$.
A CCS process $P$ may evolve to another one, say $P'$, under
  an action $\mu$, written by $P \arr{\mu} P'$.
The transition semantics of CCS processes is given by means of
a Labeled Transition System (LTS) expressed in Structural Operational
Semantics (SOS) rules shown in Fig.~\ref{f:LTSCCS}.
(The symmetric versions of the SOS rules for
parallel composition and binary choice are omitted.)
A CCS processes uses only \emph{guarded sums} if all occurrences of
sums are in the form $a.P + b.Q$. 
The \emph{immediate derivatives} of a
process $P$ are elements of the set $\{P' \st P \arr\mu P' \mbox{for some $\mu$}\}$.
We use $\ell$ to range over visible actions (i.e.~inputs or outputs, excluding  $\tau$)
and $\alpha, \mu$  to range over all actions.
\begin{figure*}[t]
\begin{center}
\vskip .1cm
 $\displaystyle{  \over  \mu.  P \arr\mu
P } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P + Q   \arr\mu
P'  } $  $ \hb$   
\hskip .5cm
 $\displaystyle{   P \arr\mu   P' \over   P | Q   \arr\mu
P' | Q } $  $ \hb$   
\hskip .3cm
  $\; \;$  $\displaystyle{ P \arr{ a}P' \hk \hk  Q
\arr{\outC a }Q'  \over  P|  Q \arr{ \tau} P'
|  Q'  }$ 
\\
\vspace{.2cm}
$\displaystyle{ P \arr{\mu}P' \over
 (\res L\!)\, P \arr{\mu} (\res L\!)\, P'} $ $ \mu, \outC\mu \not\in L$
$ \hb$
%
$\displaystyle{ P \sub {\recu A P} A \arr{\mu}P' \over
 \recu A P   \arr{ \mu} P'  } $
\hskip .5cm  
$\displaystyle{ P \arr{\mu} P' \over
 P \;[r\!f] \arr{r\!f(\mu)} P' \;[r\!f]} $
$ \hb$ %  &
\end{center}
\caption{Structural Operational Semantics of CCS}
\label{f:LTSCCS}
\end{figure*}

### 2019-09-16 ic-journal/contraction.tex
Removed all reviewed highlights in the first part of the paper; re-enable highlights

Intuitively, the bisimilarity contraction $\mcontrBIS$ is a preorder
such that, $P \mcontrBIS Q$ holds if $P \wb Q$ and, in addition, 
$Q$ \emph{has the possibility of being at least as efficient as $P$} (as far as
$\tau$-actions are performed).
Process $Q$, however, may be non-deterministic and may have other ways
of doing the same work, and these could be slow (i.e., involving
more $\tau$-\hl{actions} than those performed by $P$).
% Thus, in contrast with expansion,  we cannot really say that `$Q$ is more efficient than
% $P$'.

@@

\begin{enumerate}
\item   $P \arr\mu P'$ implies there is $Q'$ such that $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$   implies there is $P'$ such that $P \Arcap \mu
 P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\hl{Two processes $P$ and $Q$ are in the \emph{bisimilarity
    contraction}}, written as $P \mcontrBIS Q$ ($P$ \emph{contracts to} $Q$),
if $P\ \R\ Q$ for some contraction $\R$.
\hl{Sometimes we write $\mexpaBIS$ for the inverse of $\mcontrBIS$.}
\end{definition}

In the first clause $Q$ is required to match $P$'s challenge
transition with at most one transition.
This makes sure that $Q$ is capable of mimicking % verb: mimic
$P$'s work at least as efficiently as $P$. 
In contrast, the second clause of Definition~\ref{d:BisCon}, on the
challenges from $Q$, entirely ignores efficiency: it is the same
clause of  weak bisimulation~---the final derivatives are \hl{indeed} required
to be related  by $\wb$, rather than by $\R$.

Bisimilarity contraction is coarser than 
the \emph{expansion relation}
$\expa$~\cite{arun1992efficiency,sangiorgi2015equations}:
\begin{definition}%[expansion]
\label{d:expa}
A process relation ${\R}$
  is an \emph{expansion} if, whenever
we have $P\RR Q$, for all $\mu$
 \begin{enumerate}
 \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu  Q'$
  and $P' \RR Q'$;
 \item $Q \arr\mu Q'$   implies that there is $P'$ with $P \Arr \mu P'$ and $P' \RR Q'$.
 \end{enumerate}
\hl{Two processes $P$ and $Q$ are in the \emph{bisimilarity
    expansion}}, written as $P \expa Q$ ($P$ \emph{expands} $Q$),
 if $P \RR Q$ for some expansion $\R$.
 \end{definition}
The expansion preorder is widely used in proof techniques for bisimilarity and
intuitively refines bisimilarity by 
formalising the idea of ``efficiency'' between processes.
Clause (1) is the same in the \hl{both}
preorders, while in clause (2) expansion uses 
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$; 
moreover in clause (2) of Def.~\ref{d:BisCon} the final derivatives
are simply required to be bisimilar ($P' \wb Q'$).
\hl{Intuitively, $P \expa Q$ holds if $P\wb Q$ and, in addition, \emph{$Q$
  is always at least as efficient as $P$}.}

@@

\hl{In the same way as bisimilarity, both the expansion and the
contraction preorders are preserved by all CCS operators except the
summation. The proofs are similar, see e.g.}~\cite{sangiorgi2017equations}.

>>>>

Intuitively, the bisimilarity contraction $\mcontrBIS$ is a preorder
such that, $P \mcontrBIS Q$ holds if $P \wb Q$ and, in addition, 
$Q$ \emph{has the possibility of being at least as efficient as $P$} (as far as
$\tau$-actions are performed).
Process $Q$, however, may be non-deterministic and may have other ways
of doing the same work, and these could be slow (i.e., involving
more $\tau$-actions than those performed by $P$).
% Thus, in contrast with expansion,  we cannot really say that `$Q$ is more efficient than
% $P$'.

@@

\begin{enumerate}
\item   $P \arr\mu P'$ implies there is $Q'$ such that $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$   implies there is $P'$ such that $P \Arcap \mu
 P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
contraction}, written as $P \mcontrBIS Q$ ($P$ \emph{contracts to} $Q$),
if $P\ \R\ Q$ for some contraction $\R$.
Sometimes we write $\mexpaBIS$ for the inverse of $\mcontrBIS$.
\end{definition}

In the first clause $Q$ is required to match $P$'s challenge
transition with at most one transition.
This makes sure that $Q$ is capable of mimicking % verb: mimic
$P$'s work at least as efficiently as $P$. 
In contrast, the second clause of Definition~\ref{d:BisCon}, on the
challenges from $Q$, entirely ignores efficiency: it is the same
clause of  weak bisimulation~---the final derivatives are indeed required
to be related  by $\wb$, rather than by $\R$.

Bisimilarity contraction is coarser than 
the \emph{expansion relation}
$\expa$~\cite{arun1992efficiency,sangiorgi2015equations}:
\begin{definition}%[expansion]
\label{d:expa}
A process relation ${\R}$
  is an \emph{expansion} if, whenever
we have $P\RR Q$, for all $\mu$
 \begin{enumerate}
 \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu  Q'$
  and $P' \RR Q'$;
 \item $Q \arr\mu Q'$   implies that there is $P'$ with $P \Arr \mu P'$ and $P' \RR Q'$.
 \end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
  expansion}, written as $P \expa Q$ ($P$ \emph{expands} $Q$),
 if $P \RR Q$ for some expansion $\R$.
 \end{definition}
The expansion preorder is widely used in proof techniques for bisimilarity and
intuitively refines bisimilarity by 
formalising the idea of ``efficiency'' between processes.
Clause (1) is the same in the both preorders, while in clause (2) expansion uses 
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$; 
moreover in clause (2) of Def.~\ref{d:BisCon} the final derivatives
are simply required to be bisimilar ($P' \wb Q'$).
Intuitively, $P \expa Q$ holds if $P\wb Q$ and, in addition, \emph{$Q$
  is always at least as efficient as $P$}.

@@

In the same way as bisimilarity, both the expansion and the
contraction preorders are preserved by all CCS operators except the
summation. The proofs are similar, see e.g.~\cite{sangiorgi2017equations}.

### 2019-09-16 ic-journal/equa.tex
Removed all reviewed highlights in the first part of the paper; re-enable highlights

\hl{In the CCS syntax, 
a recursion $\recu A  P$ acts as a binder for $A$ in the body $P$. 
This gives rise, in the expected manner, to the notions of 
\emph{free} and \emph{bound} recursion variables in a CCS expression. 
For instance,  $X$ is free in $a.X + \recu Y (b.Y)$ while $Y$
is bound; 
in $a.X + \recu X (b.X)$,
 $X$ is
both free and bound.
%  given in previous section, there is the syntactic
% possibility that an agent variable $A$ occurs
% outside the recursion operator of the same variable (or, equivalently, a
% process constant $C$ is used but undefined). We say such a variable is
% \emph{free}, while those variables wrapped by their recursions are
% \emph{bound}.}
A tem
without free variables  is a  \emph{process}.}
% This setting does not cause any ambiguity, but sometimes leads to more
% complex proofs.} (c.f. Section~\ref{sec:multivariate} for more details.)

\hl{In this paper (and the formalisation work), we use the agent
variables also as \emph{equation variables}. This eliminates the need of
another type for  equations, and we can reuse the existing
variable substitution operation} (c.f. SOS rule for the Recursion in
Fig.~\ref{f:LTSCCS}) for the substitution of equation variables.
For example, the result of substituting variable $X$ with $\nil$ in $a.X +
\recu X (b.X)$,  written $(a.X + \recu X (b.X)) \sub {\nil} X$, is
$a.\nil + \recu X (b.X)$ (with the part $\recu X (b.X)$
untouched). \Multivariate substitutions are written in the same syntax,
e.g. $E \sub {\til P} {\til X}$. Whenever $\til X$ is clear from the
context, we may also write $E[\til P]$ instead of $E \sub {\til P} {\til
  X}$ (and $E[P]$ for $E \sub P X$ if there is a single equation variable $X$.)

% \hl{In fact, free agent variables have the same transitional behavior
% as the deadlock $\nil$ according to the SOS rules, as there is
% no rule for their transitions at all. In fact, most CCS theorems still hold if
% the involved CCS terms contain free variables. (The most notable
% exceptions are all versions of unique solution of equations/contracts,
% where all solutions must be \emph{pure} processes (i.e. no free variable).)}

\subsection{Systems of equations}
\label{ss:SysEq}
\hl{When discussing equations it is standard to talk about `context'. This is a 
 a CCS expression  possibly containing  free variables that, however, may not occur within
the body of recursive definitions. 
Milner's ``unique solution of equations'' }  theorems~\cite{Mil89} intuitively
\hl{say that, if a context $C$
%  \hl{(i.e., a CCS expression with possibly
% free variables)}\footnote{\hl{Rigorously speaking, under our setting
% (i.e.~reusing free agent variables as equation variables) not all
% valid CCS expressions
% are valid contexts: those where free variables occur inside
% recusion operators must be all excluded. For instace, $\recu X (a.X +
% b.Y)$ is not a valid context with the variable $Y$. This special requirement
% is perfectly aligned with CCS literature using process constants, as any
% equation variable cannot occur inside the definition of any
% constant. In fact, all versions of ``unique solution of
% equations/contractions'' theorems do not hold if equation variables
% are allowed to occur inside any recursion operator.}} 
obeys certain conditions,
then all processes $P$ that satisfy the equation $P \wb \ct P$ are
bisimilar with each other.
}

\begin{definition}[equations] % Def 3.1
  \label{def:equation}
Assume that, for each $i$ of 
 a countable indexing set $I$, we have variables $X_i$, and expressions
$E_i$ \hl{possibly containing such variables $\cup_i \{ X_i\}$}. Then 
$\{ X_i = E_i\}_{i\in I}$ is 
  a \emph{system of equations}. (There is one equation $E_i$ for each variable $X_i$.)
\end{definition}

@@

\begin{definition}[solutions and unique solutions]
  \label{def:solution}
Suppose $\{ X_i = E_i\}_{i\in I}$ is a system of equations: 
\begin{itemize}
\item
 $\til P$ is a \emph{solution of the system of equations (for $\wb$)} 
if for each $i$ it holds that $P_i \wb E_i [\til P]$;
\item The system has \emph{a unique solution for $\wb$}  if whenever 
 $\til P$ and $\til Q$ are both solutions \hl{then} $\til P \wb \til Q$. 
\end{itemize} 
 \end{definition}
\hl{Similarily, the \emph{(unique) solution of a system of equations for $\sim$}
(or for $\rapprox$) can be obtained by replacing all occurrences of $\wb$
in above definition with $\sim$ and $\rapprox$, respectively.}

For instance, the solution of the equation $X = a. X$ 
is the process
$R \DSdefi \recu A {\, (a. A)}$, and for any other solution $P$ we have $P \wb R$.
In contrast, the equation 
 $X = a|  X$ has solutions that may be quite different, for instance,
 $K$ and $K | b$, for $K \DSdefi \recu K {\, (a. K)}$. (Actually any process capable of
continuously performing \hl{$a$--actions} is a solution \hl{of} $X = a|  X$.)

@@

\item \hl{\emph{guarded}} if, in each $E_i$, each occurrence of
  each $X_i$ is underneath a \emph{visible} prefix;

\item \emph{sequential} if, in each $E_i$, each
  \hl{occurrence of each $X_i$ is only underneath prefixes and sums.}
\end{itemize}
\end{definition}

\hl{In other words, if a system of equations is sequential, then for
each  $E_i$, any subexpression of $E_i$ in which $X_j $
appears, apart from $X_j$ itself, is a sum of prefixed expressions.
For instance,}
\begin{itemize}
\item $X = \tau. X + \mu . \nil$ is sequential but not 
 guarded, because the guarding prefix for the variable
is not visible;
\item $X =  \ell . X | P$ is guarded but not sequential;
\item $X =  \ell . X + \tau. \res a (a .\outC b | a.\nil)$, as well as
$X = \tau . (a. X + \tau . b .X + \tau  )$ are both guarded and sequential.
\end{itemize}

\hl{Milner has  three versions of  ``unique solution of equations''
theorems, for $\sim$, $\wb$ and $\rapprox$, respectively, though only the
following two versions are explicitly mentioned} in~\citep[p.~103, 158]{Mil89}:
\begin{theorem}[unique solution of equations for $\sim$]
\label{t:Mil89s1}
Let $E_i$ be weakly guarded with free variables in $\til X$,
and let ${\til P} \sim {\til E}\{\til P /\til X\}$,
  ${\til Q} \sim {\til E}\{\til Q /\til X\}$. Then ${\til P} \sim {\til Q}$.
\end{theorem}

@@

\hl{The version of Milner's unique-solution theorem for $\wb$ further requires
that all sums are guarded}:
% \footnote{But if the CCS syntax were defined with only guarded
%   sums, i.e., $\sum_{i\in I} \mu_i.P_i$ as
%   in~\cite{sangiorgi2015equations}, this addition
%   requirement disappears automatically, and we can even say $\wb$ is indeed a
%   congruence.}:
\begin{theorem}[unique solution of equations for $\wb$]
\label{t:Mil89}
Let $E_i$ be \hl{guarded}  and sequential with free
variables up to $\til X$, and let ${\til P} \wb {\til E}\{\til P /\til X\}$,
  ${\til Q} \wb {\til E}\{\til Q /\til X\}$. Then ${\til P} \wb {\til Q}$.
\end{theorem}

>>>>

In the CCS syntax, 
a recursion $\recu A  P$ acts as a binder for $A$ in the body $P$. 
This gives rise, in the expected manner, to the notions of 
\emph{free} and \emph{bound} recursion variables in a CCS expression. 
For instance,  $X$ is free in $a.X + \recu Y (b.Y)$ while $Y$
is bound; in $a.X + \recu X (b.X)$, $X$ is both free and bound.
A \hl{term} without free variables is a \emph{process}.
% This setting does not cause any ambiguity, but sometimes leads to more
% complex proofs.} (c.f. Section~\ref{sec:multivariate} for more details.)

In this paper (and the formalisation work), we use the agent
variables also as \emph{equation variables}. This eliminates the need of
another type for  equations, and we can reuse the existing
variable substitution operation (c.f. SOS rule for the Recursion in
Fig.~\ref{f:LTSCCS}) for the substitution of equation variables.
For example, the result of substituting variable $X$ with $\nil$ in $a.X +
\recu X (b.X)$,  written $(a.X + \recu X (b.X)) \sub {\nil} X$, is
$a.\nil + \recu X (b.X)$ (with the part $\recu X (b.X)$
untouched). \Multivariate substitutions are written in the same syntax,
e.g. $E \sub {\til P} {\til X}$. Whenever $\til X$ is clear from the
context, we may also write $E[\til P]$ instead of $E \sub {\til P} {\til
  X}$ (and $E[P]$ for $E \sub P X$ if there is a single equation variable $X$.)

% In fact, free agent variables have the same transitional behavior
% as the deadlock $\nil$ according to the SOS rules, as there is
% no rule for their transitions at all. In fact, most CCS theorems still hold if
% the involved CCS terms contain free variables. (The most notable
% exceptions are all versions of unique solution of equations/contracts,
% where all solutions must be \emph{pure} processes (i.e. no free variable).)

\subsection{Systems of equations}
\label{ss:SysEq}
When discussing equations it is standard to talk about `context'. This is a 
 a CCS expression  possibly containing  free variables that, however, may not occur within
the body of recursive definitions. 
Milner's ``unique solution of equations'' theorems~\cite{Mil89} intuitively
say that, if a context $C$
%  \hl{(i.e., a CCS expression with possibly
% free variables)}\footnote{\hl{Rigorously speaking, under our setting
% (i.e.~reusing free agent variables as equation variables) not all
% valid CCS expressions
% are valid contexts: those where free variables occur inside
% recusion operators must be all excluded. For instace, $\recu X (a.X +
% b.Y)$ is not a valid context with the variable $Y$. This special requirement
% is perfectly aligned with CCS literature using process constants, as any
% equation variable cannot occur inside the definition of any
% constant. In fact, all versions of ``unique solution of
% equations/contractions'' theorems do not hold if equation variables
% are allowed to occur inside any recursion operator.}} 
obeys certain conditions,
then all processes $P$ that satisfy the equation $P \wb \ct P$ are
bisimilar with each other.

\begin{definition}[equations] % Def 3.1
  \label{def:equation}
Assume that, for each $i$ of 
 a countable indexing set $I$, we have variables $X_i$, and expressions
$E_i$ possibly containing such variables $\cup_i \{ X_i\}$. Then 
$\{ X_i = E_i\}_{i\in I}$ is 
  a \emph{system of equations}. (There is one equation $E_i$ for each variable $X_i$.)
\end{definition}

@@

\begin{definition}[solutions and unique solutions]
  \label{def:solution}
Suppose $\{ X_i = E_i\}_{i\in I}$ is a system of equations: 
\begin{itemize}
\item
 $\til P$ is a \emph{solution of the system of equations (for $\wb$)} 
if for each $i$ it holds that $P_i \wb E_i [\til P]$;
\item The system has \emph{a unique solution for $\wb$}  if whenever 
 $\til P$ and $\til Q$ are both solutions then $\til P \wb \til Q$. 
\end{itemize} 
 \end{definition}
Similarily, the \emph{(unique) solution of a system of equations for $\sim$}
(or for $\rapprox$) can be obtained by replacing all occurrences of $\wb$
in above definition with $\sim$ and $\rapprox$, respectively.

For instance, the solution of the equation $X = a. X$ 
is the process
$R \DSdefi \recu A {\, (a. A)}$, and for any other solution $P$ we have $P \wb R$.
In contrast, the equation 
 $X = a|  X$ has solutions that may be quite different, for instance,
 $K$ and $K | b$, for $K \DSdefi \recu K {\, (a. K)}$. (Actually any process capable of
continuously performing $a$--actions is a solution of $X = a|  X$.)

@@

\item \emph{guarded} if, in each $E_i$, each occurrence of
  each $X_i$ is underneath a \emph{visible} prefix;

\item \emph{sequential} if, in each $E_i$, each
  occurrence of each $X_i$ is only underneath prefixes and sums.
\end{itemize}
\end{definition}

In other words, if a system of equations is sequential, then for
each  $E_i$, any subexpression of $E_i$ in which $X_j $
appears, apart from $X_j$ itself, is a sum of prefixed expressions.
For instance,
\begin{itemize}
\item $X = \tau. X + \mu . \nil$ is sequential but not 
 guarded, because the guarding prefix for the variable
is not visible;
\item $X =  \ell . X | P$ is guarded but not sequential;
\item $X =  \ell . X + \tau. \res a (a .\outC b | a.\nil)$, as well as
$X = \tau . (a. X + \tau . b .X + \tau  )$ are both guarded and sequential.
\end{itemize}

Milner has  three versions of  ``unique solution of equations''
theorems, for $\sim$, $\wb$ and $\rapprox$, respectively, though only the
following two versions are explicitly mentioned in~\citep[p.~103, 158]{Mil89}:
\begin{theorem}[unique solution of equations for $\sim$]
\label{t:Mil89s1}
Let $E_i$ be weakly guarded with free variables in $\til X$,
and let ${\til P} \sim {\til E}\{\til P /\til X\}$,
  ${\til Q} \sim {\til E}\{\til Q /\til X\}$. Then ${\til P} \sim {\til Q}$.
\end{theorem}

@@

The version of Milner's unique-solution theorem for $\wb$ further requires
that all sums are guarded:
% \footnote{But if the CCS syntax were defined with only guarded
%   sums, i.e., $\sum_{i\in I} \mu_i.P_i$ as
%   in~\cite{sangiorgi2015equations}, this addition
%   requirement disappears automatically, and we can even say $\wb$ is indeed a
%   congruence.}:
\begin{theorem}[unique solution of equations for $\wb$]
\label{t:Mil89}
Let $E_i$ be guarded and sequential with free
variables up to $\til X$, and let ${\til P} \wb {\til E}\{\til P /\til X\}$,
  ${\til Q} \wb {\til E}\{\til Q /\til X\}$. Then ${\til P} \wb {\til Q}$.
\end{theorem}
